Ricci-Ollivier Curvature of the Rooted Phylogenetic Subtree-Prune-Regraft Graph\tnoteref{t1,t2}

### 2016-11-30 response_to_reviewers.txt

14: The phrase 'the pairings are disjoint' might need some more explanation as 'disjoint' might get interpreted in several ways (e.g. no tree is allowed to occur in more than one pairing vs. the pairings are pairwise distinct).
* revised to note "...  each neighbor of T is paired
with a single neighbor of S, and vice versa. Note that a tree may be a
neighbor of both T and S and thus involved in two different pairings."

37: The word 'and' seems to be missing before 'then' .
* corrected

Lemma 5.5:
1. Concerning the proof of (i): it might be helpful to elaborate on why a case distinction is
necessary (Why is it necessary to increase j? Why are the subcases of (not) moving the additional leaf necessary?) in order to clarify the line of argument.
* TODO

2. Concerning statement (ii): Is there a particular reason why the quotient is bounded for n ≥ 4?
The overall lemma assumes n ≥ 3, which would also lead to a nice bound (0.8).
* 5/6 is closer to the limit of 6/7 than it is to 4/5 so this gives a
	closer bound at the small cost of excluding the trivial 3-leaf tree graph.

@@

Page 21

49:

1. The meaning of the function L(.) can be guessed from the context without too much effort, but providing a short definition / explanation would help avoiding any misconceptions.
* We added "... , where L(Q) is the label set of Q, ..." to the sentence.

2. Does this equality really hold between the restricted versions of T and S , which differ only in R and, thus, should still be different after restricting them to X \ L(Q)?
* Yes, because of the triangle formed by T, S, and U, as we state in the
	proof. T and U differ by Q but not R so S is reached from U by moving Q in such a way that displaces R from its original position.

51-53: Using U as a subtree while it also denotes a whole tree in the neighbourhood of T and S just a few lines before can be slightly confusing.
* Indeed, thanks. We changed the second use of U to V and u' to v'.

@@

Page 22

30-32: The statement on the proportionality of small moves and small subtrees seems to be a bit troubled.

1. For example, if both children of the root are not small, each small subtree with one leaf contributes 2n-6 moves, while each small subtree with two leaves contributes 2n-8 moves.
So, what is the factor of proportionality?
* TODO

2. In the phrase 'where one of the root's children is small', is the intended meaning exactly one or at least one? Assuming, e.g., 2n-6 as the factor of proportionality (coming from (the majority of) the leaves), the statement does not hold in general (neither for exactly nor for at least). Consider, e.g., a tree with a leaf as the only small child of the root and at least three small subtrees with two leaves (for the case exactly) respectively the balanced tree with four leaves (for the case at least).
* TODO

36: Writing 'one 2-leaf subtree' instead of '1 2-leaf subtree' appears more readable (and is more consistent with the previous use of numerals).
* We have made the suggested change.

@@

38-42:

1. Since Lemma 5.7 shall hold for n ≥ 3, the stated number of small moves (2n2-4n-6) is not correct for all intended n. The reason is that for the ladder tree with three leaves, the small subtree with two leaves is also a child of the root and, thus, contributes 2n-3-3 (not 2n-3-5) moves. Hence, a case distinction seems to be necessary here and for the inequality in the subsequent paragraph (lines 40-42)
* TODO

2. The chain of (in)equalities in lines 40-42 does not run through, because adding (24-8n) leads to the n-free term '+12' (not '+16'). But this can be 'repaired' through the necessary case distinction n = 3 vs. n ≥ 4).
* TODO

45+55: Stating the explanation of the size '(in terms of the number of leaves)' directly in line 45 possibly avoids some confusion (which can occur when reading Lemma 5.8 while having the typical notion of size in mind).
* we added the statement to line 45.

54-55: The statement 'a ladder tree has the greatest average moved subtree size' does not seem to hold for all n ≥ 4. For n = 4, the ladder tree has an average moved subtree size of 1, while it is 4/3 for the balanced tree. Similarly, the balanced tree has a larger one for n = 8 (1.55 vs. 1.39).  This might pose a problem to the proof because it based on the idea that the ladder tree leads to themaximum.
* TODO

56: Similar to line 36, 'one leaf' and 'one subtree' appear more readable / consistent.
* We have made the suggested change.

@@

Page 23

11-13: The vertical bars denoting the size of N i (T) are missing at all three occurrences.
* corrected

11-20:

1. It is stated that |N_i(T)| ≤ 2n-3-5 for i ≤ 2, but in a ladder tree is, e.g., |N_1(T)| = (n-1)(2n-1-5) + (2n-1-3) > 2n-3-5.
* TODO

2. Also, it is stated that |N_i(T)| > 2n-3-5 for i > 2, but in a ladder tree is, e.g., |N i (T)| = 1 * 2n- (2i-1)-5 < 2n-3-5 for 2 ≤ i ≤ n-2.
* TODO

3. Therefore, also the statements on decreasing / increasing the actual values to the bound 2n-3-5 are the wrong way round. Possibly, this can be fixed by swapping the inequality signs.
* TODO

4. Is it actually obvious enough that setting all |N_i (T)| to 2n-3-5 leads to an upper for the quotient representing the average moved subtree size? A short explanation might make this point clearer.
* TODO

5. In the last sentence of the proof, the weighted average uses a proportion of 1⁄2 for both small and large subtree (which is okay), but why is the average moved subtree size of small substrees assumed to be one? Shouldn't the one small subtree with two leaves (which always exists) make it slightly larger than one?
* TODO

32: Is there are special reason for (re)defining the mass transport term
W_{1,n} ?
* W_{1,n} is used in Theorem 6.1 to distinguish between the sequential
	mass transport terms, but this this one should just be W_1. Corrected.

37: The verb 'increase' misses the third-person singular s (as it refers to 'the size of the trees').
* Corrected.

45: Only one of the two occurrences of 'thus' is needed.
* We removed the second occurence of "thus".

@@

Page 24

14: Adding a short explanation why the number of unmatched trees is in O(kn) might help the reader understanding the argument at this point better.
* We added the short explanation "There are O(kn) trees unmatched by
	this pairing, because b and a (in the nomenclature of Corollary 5.4)
	are bounded by n."

19: The innermost pair of parentheses around 'O(n)' is not necessary.
* Corrected.

30: Replacing the word 'a' before 'distance' by 'at' or 'of' conveys the meaning better.
* Replaced with "of".

31: The equality sign in the equation might be less fitting than a <, because it is previously (line 25) stated that S_n has a strictly larger neighbourhood.
* It is counterintuitive at first glance, but the equal sign is correct
	because T_n and S_n differ by an O(kn) factor.

46-47: Shortly explaining how the bounds for the distance between neighbours of T and S lead to bounds for W_1 (m_T ,m_S ) could make it easier for the reader to grasp the idea how the proof works.
* TODO

Page 25
29: Possibly it should be 'each other' instead of 'each either'.
* corrected

40-41: The actual meaning of the phrase 'a greater understanding of the distribution of tree neighborhoods with respect to one another' is not very clear (especially due to the underlined part).
* TODO

@@

Page 26

10: A short remark on how the term '1 + μ' is actually derived would make this passage easier / faster to comprehend
* TODO

14: Following Corollary 5.4 it should be 'at least γ' and not 'at most γ' paired trees (and, otherwise, the argumentation is not sensible at this point).
* Corrected.

19: The rationale for bounding the degree by using the maximum possible absolute difference (irrespective of the actual move) seems a bit unclear. Currently, it does not look like the line of argument leads to the maximum μ: using the maximum degree of T (instead of the actual, potentially smaller, value) does not result in the maximum μ because it leads to an unnecessarily large denominator.
* TODO

@@

20:

1. The last inequality ('≤ 1⁄2n2-n') holds only if n is even. For odd n, 1⁄2 has to be added to the upper bound (which is then also okay for the even case). In the end, this change also affects the curvature bound.
* TODO

2. It looks like the factors 'k' and 'a' are swapped after inserting their values. While this would not be wrong, it is a potential source of confusion.
* TODO

39-40: Providing a proof or at least a clear rationale for the statement 'the pair of trees used in the proof of Lemma 6.5 will always have negative curvature, for all n ≥ 7' would lend more weight to it.
* TODO

@@

Page 27

35: The upper bound does not seem to follow as analogously to the lower bound as it is stated.  Using the inequality stated in line 32 and proceeding analogously to the lower bound leads to κ p (T , S ) 2 ⩽ ...⩽ κ (T , S)+ p p∗max (∣ N (T )∣ , ∣ N (S )∣) where the factor p in the denominator causes a problem, because simply dropping it (in order to arrive at the bound from the lemma) bounds the value in the wrong direction (as it is at most 1).
* TODO (needs checking from the PDF)

@@

Page 28

11-12: According to Lemma 5.5, the quotient of degrees for two adjacent trees is at least 5/6 only for n ≥ 4, but there is no such assumption in Lemma 6.9 or its proof.
* TODO, add assumption

14-18: Using W 1 (T,S) (resp. W' 1 (T,S)) is slightly imprecise as T and S are the trees and not the probability densities.
* TODO

26: Within the min-term of the last case, w is the argument of all four probability densities.  Probably, two of them should have z as their argument.
* TODO

29-33: Using W 1 (resp. W' 1 ) without the arguments for the values of the Wasserstein distance is slightly imprecise.
* TODO

50: The word 'the' seems to be missing before 'understanding'.
* Corrected.

55: The statement 'rSPR graph degree changes depend quadratically on the product of the size of the regrafted subtree with its change in depth' seems a bit mixed up, because (according to the proven lemma) the changes depend linearly on the product (which in turn is quadratic in the number of leaves).
* TODO

### 2016-11-30 curvature.tex
finished first pass of revisions

\begin{lem}
	\label{lem:shared_neighbors}
Let $T$ and $S$ be trees such that $\dspr{T,S} = 1$.
Then $\size{N(T) \cap N(S)} \le 6n - 17$.
\end{lem}
\begin{proof}
	$T$ and $S$ differ by one rSPR move that moves a subtree $R$.
	Pick a neighbor $U \in N(T) \cap N(S)$ of both $T$ and $S$ (this intersection is not empty: $T$ and $S$ are different, so $R$ contains at most $n-2$ of the leaves, thus there must be at least one other tree $U$ obtained by moving $R$ in $T$ and $S$).
	Then either (i) $T$ and $U$ differ in the location of $R$, or (ii) $T$ and $U$ differ in the location of another subtree $Q$.
	In the latter case, $T|(X \setminus L(Q)) = S|(X \setminus L(Q))$ because $T$ and $S$ differ only in the location of $R$ and $\dspr{T,U} = \dspr{S,U} = 1$.
	Then leaves $r' \in R$, $q' \in Q$, and $u' \in U$, for some subtree $U$, form a triple of $T$ and a different triple in $S$.
	This incompatible triple can be resolved in at most $6n - 17$ ways, the maximum of which is reached when $Q$, $U$, and $R$ are themselves a ``triple'' of subtrees.
	By Lemma~\ref{lem:compute_degree}, each of the subtrees is assigned to at most $2n-6$ unique moves.
	Moreover, one additional overlapping move also moves one of the subtrees (that of the aunt of the LCA of the three subtrees).
	The number of shared neighbors is thus at most $3(2n-6) + 1 = 6n-17$.
	Note that this bound is tight when, for example, $T$ and $S$ are ladders with a different configuration of 3 leaves at maximum depth.
\end{proof}

@@

	A ladder tree has the smallest proportion of small subtrees, so we first bound the number of small subtree moves that can be applied to a ladder tree.
	A ladder tree with $n$ leaves has $n$ 1-leaf subtrees and 1 2-leaf subtree.
	One of the 1-leaf subtrees is a child of the root.
	A 2-leaf subtree contains three nodes.
	Therefore, a ladder tree induces $n (2n - 1 - 5) + 2 + (2n - 3 - 5) = 2n^2 - 4n - 6$ small subtree moves.

@@

Now, we bound the average size of the moved subtrees.

\begin{lem}
	Let $T$ be a rooted tree with $n \ge 4$ leaves.
	Let $N_i(T)$ be the neighbors of $T$ induced by rSPR moves of an $i$-leaf subtree.
	Then $$\frac{\sum_i i \size{N_i(T)}}{\size{N(T)}} \le \frac{n + 3}{4} .$$
	\label{lem:spr_size_mean}
\end{lem}
\begin{proof}
	Again recall our distribution of rSPR moves from a tree in Lemma~\ref{lem:compute_degree}.
	By similar arguments as in the proof of Lemma~\ref{lem:spr_size}, a ladder tree has the greatest average moved subtree size (in terms of the number of leaves).
	There are $n$ subtrees with 1 leaf, and 1 subtree each with $2,3, \ldots n - 1$ leaves.
	We ignore the root subtree, which does not induce any rSPR moves.
	Thinking of the ratio as a weighted average of $i$, we can split the numerator and the denominator of the average into terms with $i \leq 2$ and those with $i > 2$.
    $N_i(T) \leq 2n - 3 - 5$ for the first set of terms, and $N_i(T) > 2n - 3 - 5$ for the second, so by replacing $N_i(T)$ in each case by $2n-3-5$ we decrease the weight for $i \leq 2$ and increase the weight on $i > 2$.
	Thus, by doing this replacement we obtain a upper bound on the mean moved subtree size.
	This bound, ignoring small subtrees, is $(3 + (n-2)) / 2 = (n + 1) / 2$.
	By Lemma~\ref{lem:spr_size}, at least half of the rSPR moves involve a small subtree.
	Thus, the average size of a moved subtree over all SPR operations from a given tree is at most the weighted average of $1 / 2 + ((n + 1) / 2) / 2$ = $(n + 1) / 4 + 1 / 2 = (n + 3) / 4$.

@@

Recall that the uniform random walk begins at a tree $T$ and moves to a tree uniformly at random from $N(T)$.
Recall that the coarse uniform random walk curvature between two trees $T$ and $S$ is $\curvature{T, S} := 1 - \frac{W_1(m_T, m_S)}{d(T, S)}$, where $W_{1,n}$ is the mass transport term \eqref{eq:curvatureDef}.
For the uniform random walk, $m_T$ is the probability measure assigning a mass of $\frac{1}{\degree{T}}$ to each of $T$'s neighbors.

We showed in Lemmas~\ref{lem:spr_size} and~\ref{lem:spr_size_mean} that most adjacent trees differ by a small subtree, raising the question of what the curvature is between such pairs.
In our first theorem of this section, we show that the curvature between two adjacent trees tends towards 0 as the size of the trees increase but the size of the subtree moved to transition between the two trees remains the same size.
To do so, we prove that the mass transport term between adjacent trees $T$ and $S$ is dominated by probability moved between neighbors of $T$ and $S$ that are also adjacent.
The probability transported between members of these squares (Corollary~\ref{cor:paired_neighbors}) increases more quickly with $n$ than the probability transported between shared neighbor triangles (Lemma~\ref{lem:shared_neighbors}), and outweighs the effect of long distance probability moved between more distant neighbors or due to a difference in the degree of $T$ and $S$.
Thus this ``flatness-in-the-limit'' theorem thus helps explain our empirical results in Section~\ref{sec:results:curvature} showing that the distribution of curvature pairs tends more towards 0 as $n$ increases.

@@

To start we demonstrate the theorem in the case that $T_n$ and $S_n$ have the same number of neighbors.
First we claim that $W_{1,n}$ is bounded above by $(\degree{T_n}+\OhOf{kn})/\degree{T_n}$ by exhibiting a mass transport program satisfying that bound.
Let $(T'_n, S'_n)$ be any of the $\overlap$ pairs of neighbors of $(T_n, S_n)$ which are one rSPR move apart as per Corollary~\ref{cor:paired_neighbors}.
We pair these trees in the mass transport.
There are $\OhOf{kn}$ trees unmatched by this pairing, and we can pair each of them arbitrarily with another tree of distance at most 3.
Thus, $W_{1,n}$ is bounded above by $(\degree{T_n} + \OhOf{kn}) / \degree{T_n}$.

A lower bound is also available because we can't do better than distance 1 for all trees except for shared neighbors, of which there are $\OhOf{n}$ by Lemma~\ref{lem:shared_neighbors}.
By ignoring these trees we get a lower bound of $(\degree{T_n} - (\OhOf{n}))/\degree{T_n}$ for $W_{1,n}$.

@@

Now we prove the theorem when the number of neighbors differ.
Assume without loss of generality that $\degree{T_n} < \degree{S_n}$.
By Lemma~\ref{lem:degree_change}, $\degree{S_n} - \degree{T_n} = 2(k(a-b) + i - j)$, where each of $\{a,b,i,j\}$ is less than $n$.
Thus, $\degree{S_n} - \degree{T_n} = \OhOf{kn}$.
We again pair neighbor $T'_n$ of $T$ with neighbor $S'_n$ of $S$ such that $\dspr{T'_n, S'_n}=1$ but, as $\degree{T_n} < \degree{S_n}$ we can only account for at most $\degree{T_n} / \degree{S_n}$ of the mass directly and may have to move the $(\degree{S_n} - \degree{T_n}) / \degree{S_n}$ remainder to trees a distance at most 3.
Thus, $W_{1,n}$ is bounded above by $(\degree{T_n} + \OhOf{kn}) / \degree{S_n} = (\degree{S_n} + \OhOf{kn}) / \degree{S_n}$.
We again bound $W_{1,n}$ from below with $(\degree{T_n} - \OhOf{n}) / \degree{T_n}$ by ignoring the mass in common neighbors of $T_n$ and $S_n$.
The theorem again follows because $\degree{T_n}$ is quadratic in $n$.
\end{proof}

@@

It is more difficult to obtain a closer bound on the maximum curvature of nonadjacent trees.
Lemma~\ref{lem:curvature_distance_bound} suggests that more distant pairs of trees should have smaller curvatures than close trees, because neighborhood effects decrease with respect to the increasing distance.
However, our experiments with $n \le 7$ suggest that maximum curvature tends to increase with distance (with respect to a fixed $n$), as a far greater fraction of the neighbors approach each other as the distance increases.
Indeed, for $5 \le n \le 7$ the maximum curvature is obtained by pairs of trees at one less than the maximum distance.
Moreover, nearly all of the neighbors of these pairs approach each either.
We thus conjecture the following:
\begin{conj}
	Let $k_n$ be the maximum curvature between two trees with $n$-leaves.
	Then:
	\begin{enumerate}
		\item $k_n \le \frac{2}{\Delta_{\text{rSPR}}(n)-1}$, and
		\item $ k_n \sim \frac{2}{\Delta_{\text{rSPR}}(n)-1}$.
	\end{enumerate}
\end{conj}
Proving or disproving this conjecture would go a long way toward understanding the effects of relative distance on curvature.
However, we suspect that this will require a greater understanding of the distribution of tree neighborhoods with respect to one another than is currently known.

@@

	By Lemmas~\ref{lem:degree_change} and~\ref{lem:degree_max_delta_adjacent}, the latter effect is maximized when the relative degree change is maximized.
	By Corollary~\ref{cor:paired_neighbors}, there are at most $\overlap := \degree{T} -2ka - 2(i-1)$ paired trees, bounding the former effect.
	We now construct a pair of trees that maximizes both effects.
	Let $S$ be the ladder tree with degree $3n^2 - 13n + 14$ and $T$ be the adjacent tree constructed by moving the lower $\floor{\frac{n}{2}}$ leaves of $S$ to the root.
	$T$ has degree at most $3.5n^2 -15n + 16$.
	There are thus $2ka + 2(i-1) = 2\left(\ceil{\frac{n-2}{2}}\floor{\frac{n}{2}} +(1-1)\right) \le \frac{1}{2}n^2 - n$ unpaired neighbors, the maximum possible.
	Moreover, as shown by Lemma~\ref{lem:degree_change} this pair of trees obtains the maximum (absolute and relative) degree change.
	Thus, the maximum $\mu$ is:
	$$\frac{\frac{1}{2}n^2 - n}{3.5n^2 - 15n + 16}.$$
	The claim follows from multiplying this value by $-2$.
\end{proof}

@@

\section{Conclusion and future work}
In summary, we have gone beyond graph diameter and vertex degree to substantially advance understanding of the phylogenetic rSPR graph.
We did so by developing the first theoretical and computational frameworks to bound and compute Ricci-Ollivier curvature of the rSPR graph.
We found that curvature, along with degree and distance, determine the early dynamics of hitting times for random walks.
Moreover, we proved that rSPR graph degree changes depend quadratically on the product of the size of the regrafted subtree with its change in depth.
As a result of this dependence, some areas of the rSPR graph are locally negatively curved but the graph tends toward flatness with respect to the majority of rSPR moves which move small subtrees.
Finally, we proved that the coarse and asymptotic definitions of Ricci-Ollivier curvature are closely related with respect to uniform and Metropolis-Hastings walks on the rSPR graph.

In this data-free setting the stationary distribution is, unlike with real data, quite evenly spread over all trees.
Correspondingly, we found that the influence of curvature is small in this case (Fig.~\ref{fig:kappa-access-distribution}) and that the probability of the target node in the stationary distribution predominantly determines access times for pairs of trees (Fig.~\ref{fig:kappa-access}).
However, it is well known that MCMC takes a long time to approximate real phylogenetic posterior distributions even when the Bayesian credible set is small, and in fact our previous work showed significant SPR graph influence on the mixing time for phylogenetic MCMC for credible sets that had tens, hundreds or thousands of trees~\citep{Whidden2015-yi}.
Thus, our next step will be to investigate curvature of MCMC with nontrivial likelihood functions, which will reduce the posterior distribution to a more realistic effective size, and in certain cases will lead to significant ``bottlenecks'' like those we have observed in real data.
In those cases the curvature between two trees at either end of a bottleneck will describe how difficult it is to traverse the bottleneck.
Indeed, in both the setting of challenging real data and in simulation with a nontrivial likelihood function, the stationary distribution takes a long time to achieve, and thus the curvature will have a substantially greater impact on the overall hitting times rather than being a short prelude to waiting for a Poisson process event as it is here (Fig.~\ref{fig:kappa-access}).
One avenue of exploration will be to follow previous work \citep{Stefankovic2011-hu} by taking a chain where tree topology likelihoods are maximized across branch length, as the classical phylogenetic likelihood \citep{felsenstein1981evolutionary} requires both branch length and topology.

>>>>

\begin{lem}
	\label{lem:shared_neighbors}
Let $T$ and $S$ be trees such that $\dspr{T,S} = 1$.
Then $\size{N(T) \cap N(S)} \le 6n - 17$.
\end{lem}
\begin{proof}
	$T$ and $S$ differ by one rSPR move that moves a subtree $R$.
	Pick a neighbor $U \in N(T) \cap N(S)$ of both $T$ and $S$ (this intersection is not empty: $T$ and $S$ are different, so $R$ contains at most $n-2$ of the leaves, thus there must be at least one other tree $U$ obtained by moving $R$ in $T$ and $S$).
	Then either (i) $T$ and $U$ differ in the location of $R$, or (ii) $T$ and $U$ differ in the location of another subtree $Q$.
	In the latter case, $T|(X \setminus L(Q)) = S|(X \setminus L(Q))$, where $L(Q)$ is the label set of $Q$, because $T$ and $S$ differ only in the location of $R$ and $\dspr{T,U} = \dspr{S,U} = 1$.
	Then leaves $r' \in R$, $q' \in Q$, and $v' \in V$, for some subtree $V$, form a triple of $T$ and a different triple in $S$.
	This incompatible triple can be resolved in at most $6n - 17$ ways, the maximum of which is reached when $Q$, $U$, and $R$ are themselves a ``triple'' of subtrees.
	By Lemma~\ref{lem:compute_degree}, each of the subtrees is assigned to at most $2n-6$ unique moves.
	Moreover, one additional overlapping move also moves one of the subtrees (that of the aunt of the LCA of the three subtrees).
	The number of shared neighbors is thus at most $3(2n-6) + 1 = 6n-17$.
	Note that this bound is tight when, for example, $T$ and $S$ are ladders with a different configuration of 3 leaves at maximum depth.
\end{proof}

@@

	A ladder tree has the smallest proportion of small subtrees, so we first bound the number of small subtree moves that can be applied to a ladder tree.
	A ladder tree with $n$ leaves has $n$ 1-leaf subtrees and one 2-leaf subtree.
	One of the 1-leaf subtrees is a child of the root.
	A 2-leaf subtree contains three nodes.
	Therefore, a ladder tree induces $n (2n - 1 - 5) + 2 + (2n - 3 - 5) = 2n^2 - 4n - 6$ small subtree moves.

@@

Now, we bound the average size of the moved subtrees (in terms of the number of leaves).

\begin{lem}
	Let $T$ be a rooted tree with $n \ge 4$ leaves.
	Let $N_i(T)$ be the neighbors of $T$ induced by rSPR moves of an $i$-leaf subtree.
	Then $$\frac{\sum_i i \size{N_i(T)}}{\size{N(T)}} \le \frac{n + 3}{4} .$$
	\label{lem:spr_size_mean}
\end{lem}
\begin{proof}
	Again recall our distribution of rSPR moves from a tree in Lemma~\ref{lem:compute_degree}.
	By similar arguments as in the proof of Lemma~\ref{lem:spr_size}, a ladder tree has the greatest average moved subtree size (in terms of the number of leaves).
	There are $n$ subtrees with one leaf, and one subtree each with $2,3, \ldots n - 1$ leaves.
	We ignore the root subtree, which does not induce any rSPR moves.
	Thinking of the ratio as a weighted average of $i$, we can split the numerator and the denominator of the average into terms with $i \leq 2$ and those with $i > 2$.
    $\size{N_i(T)} \leq 2n - 3 - 5$ for the first set of terms, and $\size{N_i(T)} > 2n - 3 - 5$ for the second, so by replacing $\size{N_i(T)}$ in each case by $2n-3-5$ we decrease the weight for $i \leq 2$ and increase the weight on $i > 2$.
	Thus, by doing this replacement we obtain a upper bound on the mean moved subtree size.
	This bound, ignoring small subtrees, is $(3 + (n-2)) / 2 = (n + 1) / 2$.
	By Lemma~\ref{lem:spr_size}, at least half of the rSPR moves involve a small subtree.
	Thus, the average size of a moved subtree over all SPR operations from a given tree is at most the weighted average of $1 / 2 + ((n + 1) / 2) / 2$ = $(n + 1) / 4 + 1 / 2 = (n + 3) / 4$.

@@

Recall that the uniform random walk begins at a tree $T$ and moves to a tree uniformly at random from $N(T)$.
Recall that the coarse uniform random walk curvature between two trees $T$ and $S$ is $\curvature{T, S} := 1 - \frac{W_1(m_T, m_S)}{d(T, S)}$, where $W_1$ is the mass transport term \eqref{eq:curvatureDef}.
For the uniform random walk, $m_T$ is the probability measure assigning a mass of $\frac{1}{\degree{T}}$ to each of $T$'s neighbors.

We showed in Lemmas~\ref{lem:spr_size} and~\ref{lem:spr_size_mean} that most adjacent trees differ by a small subtree, raising the question of what the curvature is between such pairs.
In our first theorem of this section, we show that the curvature between two adjacent trees tends towards 0 as the size of the trees increases but the size of the subtree moved to transition between the two trees remains the same size.
To do so, we prove that the mass transport term between adjacent trees $T$ and $S$ is dominated by probability moved between neighbors of $T$ and $S$ that are also adjacent.
The probability transported between members of these squares (Corollary~\ref{cor:paired_neighbors}) increases more quickly with $n$ than the probability transported between shared neighbor triangles (Lemma~\ref{lem:shared_neighbors}), and outweighs the effect of long distance probability moved between more distant neighbors or due to a difference in the degree of $T$ and $S$.
Thus this ``flatness-in-the-limit'' theorem helps explain our empirical results in Section~\ref{sec:results:curvature} showing that the distribution of curvature pairs tends more towards 0 as $n$ increases.

@@

To start we demonstrate the theorem in the case that $T_n$ and $S_n$ have the same number of neighbors.
First we claim that $W_{1,n}$ is bounded above by $(\degree{T_n}+\OhOf{kn})/\degree{T_n}$ by exhibiting a mass transport program satisfying that bound.
Let $(T'_n, S'_n)$ be any of the $\overlap$ pairs of neighbors of $(T_n, S_n)$ which are one rSPR move apart as per Corollary~\ref{cor:paired_neighbors}.
We pair these trees in the mass transport.
There are $\OhOf{kn}$ trees unmatched by this pairing, because $b$ and $a$ (in the nomenclature of Corollary~\ref{cor:paired_neighbors}) are bounded by $n$.
We can pair each of the unmatched trees arbitrarily with another tree of distance at most 3.
Thus, $W_{1,n}$ is bounded above by $(\degree{T_n} + \OhOf{kn}) / \degree{T_n}$.

A lower bound is also available because we can't do better than distance 1 for all trees except for shared neighbors, of which there are $\OhOf{n}$ by Lemma~\ref{lem:shared_neighbors}.
By ignoring these trees we get a lower bound of $(\degree{T_n} - \OhOf{n})/\degree{T_n}$ for $W_{1,n}$.

@@

Now we prove the theorem when the number of neighbors differ.
Assume without loss of generality that $\degree{T_n} < \degree{S_n}$.
By Lemma~\ref{lem:degree_change}, $\degree{S_n} - \degree{T_n} = 2(k(a-b) + i - j)$, where each of $\{a,b,i,j\}$ is less than $n$.
Thus, $\degree{S_n} - \degree{T_n} = \OhOf{kn}$.
We again pair neighbor $T'_n$ of $T$ with neighbor $S'_n$ of $S$ such that $\dspr{T'_n, S'_n}=1$ but, as $\degree{T_n} < \degree{S_n}$ we can only account for at most $\degree{T_n} / \degree{S_n}$ of the mass directly and may have to move the $(\degree{S_n} - \degree{T_n}) / \degree{S_n}$ remainder to trees of distance at most 3.
Thus, $W_{1,n}$ is bounded above by $(\degree{T_n} + \OhOf{kn}) / \degree{S_n} = (\degree{S_n} + \OhOf{kn}) / \degree{S_n}$.
We again bound $W_{1,n}$ from below with $(\degree{T_n} - \OhOf{n}) / \degree{T_n}$ by ignoring the mass in common neighbors of $T_n$ and $S_n$.
The theorem again follows because $\degree{T_n}$ is quadratic in $n$.
\end{proof}

@@

It is more difficult to obtain a closer bound on the maximum curvature of nonadjacent trees.
Lemma~\ref{lem:curvature_distance_bound} suggests that more distant pairs of trees should have smaller curvatures than close trees, because neighborhood effects decrease with respect to the increasing distance.
However, our experiments with $n \le 7$ suggest that maximum curvature tends to increase with distance (with respect to a fixed $n$), as a far greater fraction of the neighbors approach each other as the distance increases.
Indeed, for $5 \le n \le 7$ the maximum curvature is obtained by pairs of trees at one less than the maximum distance.
Moreover, nearly all of the neighbors of these pairs approach each other.
We thus conjecture the following:
\begin{conj}
	Let $k_n$ be the maximum curvature between two trees with $n$-leaves.
	Then:
	\begin{enumerate}
		\item $k_n \le \frac{2}{\Delta_{\text{rSPR}}(n)-1}$, and
		\item $ k_n \sim \frac{2}{\Delta_{\text{rSPR}}(n)-1}$.
	\end{enumerate}
\end{conj}
Proving or disproving this conjecture would go a long way toward understanding the effects of relative distance on curvature.
However, we suspect that this will require a greater understanding of the distribution of tree neighborhoods with respect to one another than is currently known.

@@

	By Lemmas~\ref{lem:degree_change} and~\ref{lem:degree_max_delta_adjacent}, the latter effect is maximized when the relative degree change is maximized.
	By Corollary~\ref{cor:paired_neighbors}, there are at least $\overlap := \degree{T} -2ka - 2(i-1)$ paired trees, bounding the former effect.
	We now construct a pair of trees that maximizes both effects.
	Let $S$ be the ladder tree with degree $3n^2 - 13n + 14$ and $T$ be the adjacent tree constructed by moving the lower $\floor{\frac{n}{2}}$ leaves of $S$ to the root.
	$T$ has degree at most $3.5n^2 -15n + 16$.
	There are thus $2ka + 2(i-1) = 2\left(\ceil{\frac{n-2}{2}}\floor{\frac{n}{2}} +(1-1)\right) \le \frac{1}{2}n^2 - n$ unpaired neighbors, the maximum possible.
	Moreover, as shown by Lemma~\ref{lem:degree_change} this pair of trees obtains the maximum (absolute and relative) degree change.
	Thus, the maximum $\mu$ is:
	$$\frac{\frac{1}{2}n^2 - n}{3.5n^2 - 15n + 16}.$$
	The claim follows from multiplying this value by $-2$.
\end{proof}

@@

\section{Conclusion and future work}
In summary, we have gone beyond graph diameter and vertex degree to substantially advance the understanding of the phylogenetic rSPR graph.
We did so by developing the first theoretical and computational frameworks to bound and compute Ricci-Ollivier curvature of the rSPR graph.
We found that curvature, along with degree and distance, determine the early dynamics of hitting times for random walks.
Moreover, we proved that rSPR graph degree changes depend quadratically on the product of the size of the regrafted subtree with its change in depth.
As a result of this dependence, some areas of the rSPR graph are locally negatively curved but the graph tends toward flatness with respect to the majority of rSPR moves which move small subtrees.
Finally, we proved that the coarse and asymptotic definitions of Ricci-Ollivier curvature are closely related with respect to uniform and Metropolis-Hastings walks on the rSPR graph.

In this data-free setting the stationary distribution is, unlike with real data, quite evenly spread over all trees.
Correspondingly, we found that the influence of curvature is small in this case (Fig.~\ref{fig:kappa-access-distribution}) and that the probability of the target node in the stationary distribution predominantly determines access times for pairs of trees (Fig.~\ref{fig:kappa-access}).
However, it is well known that MCMC takes a long time to approximate real phylogenetic posterior distributions even when the Bayesian credible set is small, and in fact our previous work showed significant SPR graph influence on the mixing time for phylogenetic MCMC for credible sets that had tens, hundreds or thousands of trees~\citep{Whidden2015-yi}.
Thus, our next step will be to investigate curvature of MCMC with nontrivial likelihood functions, which will reduce the posterior distribution to a more realistic effective size, and in certain cases will lead to significant ``bottlenecks'' like those we have observed in real data.
In those cases the curvature between two trees at either end of a bottleneck will describe how difficult it is to traverse the bottleneck.
Indeed, in both the setting of challenging real data and in simulation with a nontrivial likelihood function, the stationary distribution takes a long time to be achieved, and thus the curvature will have a substantially greater impact on the overall hitting times rather than being a short prelude to waiting for a Poisson process event as it is here (Fig.~\ref{fig:kappa-access}).
One avenue of exploration will be to follow previous work \citep{Stefankovic2011-hu} by taking a chain where tree topology likelihoods are maximized across branch length, as the classical phylogenetic likelihood \citep{felsenstein1981evolutionary} requires both branch length and topology.
